Analysis of 'Step' Data
==================================

## Introduction

This document provides an analysis of 'Step' data submitted by an anonymous individual who collected 2 months worth of data, from a device that counted steps in 5 minute intervals.

A copy of the original dataset is available [here](
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

## Information about the data 

The data is organized simply as "steps,date,interval", where:

* **steps** = number steps taken in a 5 minute interval
* **date** = date of collection
* **interval** = indicates the interval of the day by counting collective minutes after midnight

### Load the data

```{r loaddata}
stepdata <- read.csv("activity.csv", header = TRUE)
str(stepdata)
```


### Total steps taken

First, we would like to look at a histogram showing the total number steps taken:

```{r dailysum}
# count steps
dailysum <- tapply(stepdata$steps, stepdata$date, FUN=sum, na.rm=TRUE)
```
```{r histdailysum,echo=FALSE}
hist(dailysum, breaks="FD", xlab="Total Step Count", main="Frequency of Total Step Count")
```


### Mean and Median

Looking further at the data, the mean and median of total steps taken per day are:

```{r meanmedian}
values <- unname(dailysum)
mean_ds <- mean(sort(values))
med_ds <- median(sort(values))
```
```{r showmean}
# mean
mean_ds

#median
med_ds
```


### Time Series of steps taken by time interval

The number of steps by time interval can also be looked at:

```{r timeseries}
ts <- aggregate(stepdata$steps, list(stepdata$interval), mean, na.rm=TRUE)
names(ts) <- c("interval","mean")
```
```{r plottimeseries, echo=FALSE}
plot(ts, type="l", main="Average Steps by Time Interval", xlab="Time Interval", ylab="Step Count")
```


### Interval with highest average step count

Calculating the interval with the highest average steps:

```{r maxtimeseries}
max(ts$mean)
maxinterval <- ts[which(ts$mean == max(ts$mean)),1]
maxinterval
```


### Re-analyze the Data, accounting for 'NA'

Re-calculate the above analysis, making sure all 'NA's have been accounted for.  Where an 'NA' is encountered, we will replace the NA with the mean of the interval time series.

Summary of NA values:
```{r countna}
colSums(is.na(stepdata))
```


Adjust the step data, to replace all NA values with the mean from timeseries calculation:

```{r indexna}

adjusteddata <- stepdata

for(i in 1:length(adjusteddata[,1])){

    if (is.na(adjusteddata$steps[i])) {
        inter <- adjusteddata$interval[i]
        newmean <- ts$mean[ts$interval == inter]
        adjusteddata$steps[i] <- newmean
    }
}

head(adjusteddata)
```


### Total steps taken - Adjusted

Recalculate total number steps taken:

```{r adjusteddailysum}
# count steps
adjusteddailysum <- tapply(adjusteddata$steps, adjusteddata$date, FUN=sum)
```
```{r plotadjusteddailysum, echo=FALSE}
hist(adjusteddailysum, breaks="FD", xlab="Total Step Count", main="Frequency of Total Step Count")
```


### Mean and Median - Adjusted

Adjusted mean and median of total steps taken per day are:

```{r adjustedmeanmedian}
adjvalues <- unname(adjusteddailysum)

adj_mean_ds <- mean(adjvalues)
adj_med_ds <- median(adjvalues)
```

```{r showadjmean}
# adjusted mean
adj_mean_ds

# adjusted median
adj_med_ds
```


### Analysis of Weekday vs. Weekend Activity


```{r addweekdayfactor}
dayvector <- vector(length=length(adjusteddata$date))

for(i in 1:length(adjusteddata[,1])){

    dayvector[i] <- 
        
        ifelse( weekdays(as.POSIXlt(stepdata$date[i])) == "Sunday", "Weekend", 

            ifelse( weekdays(as.POSIXlt(stepdata$date[i])) == "Saturday", "Weekend", "Weekday")
        )
}

adjusteddata$day <- dayvector

str(adjusteddata)
    
```


### Comparing Weekday vs. Weekend Activity

```{r panelplot,echo=FALSE}
par(mfrow = c(1,2))
plot(steps ~ interval, data=subset(adjusteddata, adjusteddata$day == "Weekday"), main="Weekdays", type="l")
plot(steps ~ interval, data=subset(adjusteddata, adjusteddata$day == "Weekend"), main="Weekends", type="l")
```


